Liveness-Based Garbage Collection for Lazy Languages

### 2015-07-10 paper/paper.tex
typo fixes.

{\color {Myblue}Rewrite  after the  experiments: Experiments
  with  a single  generation  copying collector  confirm the  expected
  performance  benefits.   Liveness-based  collection  results  in  an
  increase  in  garbage  reclaimed.   As a  consequence,  there  is  a
  decrease in  the number of  collections and a  (some quantification)
  decrease in the memory size  required to run programs. Besides there
  is a reduction in the overall  execution time for some programs. Our
  experiments further show that a mixed strategy, where we do liveness
  based GC  up to the closure  boundary and reachability based  GC for
  the  arguments (free  variables) of  the closure,  gives us  similar
  benefits as  a fully liveness based  GC (where the arguments  of the
  closure are also collected using liveness) at much less overhead (?).
}

%------------------------------------------------------------%
\subsection{Motivating Example}
\label{sec:motiv}
\begin{figure}[t!]
  \psset{unit=1mm}
  \begin{pspicture}(90,48)(0,-48)
    %\psframe(90,48)(0,-48)
    \begin{tabular}{@{}c@{}}
      %\psframebox
          {\sf
	\renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
            \UFL\
            \UNL{0} (\DEFINE\ (\length\  \pl)
	    \UNL{1}  (\SIF~(\NULLQ \ \pl) $0$
            (\PRIM\ 1\ (\CAR\  \pl))))
            \UNL{0}
	    \UNL{0}  (\DEFINE\ (\append\  \lista\ \listb)
	    \UNL{1}  (\SIF~(\NULLQ \ \lista)
	    \listb
	    \UNL{2} \hspace*{-0.1cm}(\CONS\ (\CAR\  \lista) (\append\
            (\CDR\  \lista)
            \listb))))
            \UNL{0}
            \UNL{0} (\LET\ \px\
            $\leftarrow$\ (\CONS\ $5$
            (\CONS\ (\CONS\ $6$ \NIL) \NIL) \IN
	    \UNL{1} (\LET\ \py\   $\leftarrow$\  (\CONS\ $3$ \NIL) \IN
	    \UNL{2}
            (\LET\ \pz\  $\leftarrow$\  (\append\ \px\  \py)\ \IN\
            \UNL{3} (\SIF~(\NULLQ~(\CAR~\pz))~$0$~$\pi$:\
            (\length\ \pz))))))
	  \end{uprogram}
      }}
      \\ 
      (a) Example program. \\  \\
%      \includegraphics[width=.45\textwidth]{motiv-example}
      \input{mem-graph} \\ 
      \renewcommand{\arraystretch}{.9}
      \begin{tabular}[t]{p{0.9\columnwidth}}
        (b) Memory graph at $\pi$.  \scalebox{.7}{\TwoCellsAD{a1}{a2}}
        denotes a closure. Thick edges denote live links. Traversal
        stops at edges marked $\times$ during garbage collection.
      \end{tabular}
    \end{tabular} 
  \end{pspicture}
  \vspace*{-3ex}
  \caption{Example Program and its Memory Graph}\label{fig:mot-example}
\end{figure}

@@

The edges shown by thick arrows are  those which are live at $\pi$.  A
cell is  preserved during garbage  collection only if it  is reachable
from the root set  through a path of live edges.   All other cells can
be reclaimed.  Thus if a garbage  collection takes place at $\pi$ with
the  heap shown  in Figure~\ref{fig:mot-example}(b),  a liveness-based
collector  (LGC) will  preserve only  the cells  referenced by  $\pz$,
$(\CDR~ \pz)$ and  (the cells constituting) the  closure referenced by
$(\CDR~(\CDR~\pz))$.   In  contrast,  a  reachability-based  collector
(RGC) would have preserved all cells.

@@

{
\color {Myblue}[I don't think this is valid anymore!!]
Note that  liveness refers to  access using variables and  fields, and
not  to $\CONS$  cells (i.e.\  to edges  in the  memory graph,  not to
locations themselves).
}

@@

 Since  in a  lazy language,  an  expression is  not evaluated  unless
 required, the null demand ($\emptyset$) does not generate liveness in
 any  of  the  rules  defining  $\mathit{ref}$  or  $\mathcal{L}$.   A
 non-null  demand of  $\sigma$ on  (\CDR~$x$), is  transformed  to the
 liveness $\{x.\epsilon, x.\acdr\sigma\}$.   In an opposite sense, the
 demand  of $\acdr\sigma$  on  (\CONS~$y$~$z$) is  transformed to  the
 demand  $\sigma$ on  $z$.   Note  that \CONS\  does  not, by  itself,
 dereference its  arguments. The rules for  (\PRIM~x~y) and (\NULLQ~x)
 are similar. Constants do not generate any liveness.

@@

Section~\ref{sec:computing}   shows   how   the  demand   transformers
$\Lfonly$  for  a  program  (representing  a  fully  context-sensitive
analysis) can be safely approximated by a {\em procedure summary}. for
each function.   The summary  is in the  form of a  demand transformer
that maps a demand on a call to the function to demands on each of its
arguments.

>>>>

{\color  {Myblue}   Experiments  with  a  single   generation  copying
  collector confirm the expected performance benefits.  Liveness-based
  collection  results  in an  increase  in  garbage reclaimed.   As  a
  consequence, there is a decrease in  the number of collections and a
  (some quantification)  decrease in the  memory size required  to run
  programs. Besides there is a reduction in the overall execution time
  for  some  programs.  Our  experiments  further  show that  a  mixed
  strategy, where we  do liveness based GC up to  the closure boundary
  and reachability based GC for  the arguments (free variables) of the
  closure,  gives us  similar benefits  as a  fully liveness  based GC
  (where  the  arguments  of  the closure  are  also  collected  using
  liveness)  at  much  less  overhead (?).  [Overhead:  extra  memory,
    book-keeping during run-time to update, multiple traversal of same
    set of cells][Refer to Section~\ref{some-sec-in-impl}] }

%------------------------------------------------------------%
\subsection{Motivating Example}
\label{sec:motiv}
\begin{figure}[t!]
  \psset{unit=1mm}
  \begin{pspicture}(90,48)(0,-48)
    %\psframe(90,48)(0,-48)
    \begin{tabular}{@{}c@{}}
      %\psframebox
          {\sf
	\renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
            \UFL\
            \UNL{0} (\DEFINE\ (\length\  \pl)
	    \UNL{1}  (\SIF~(\NULLQ \ \pl) $0$
            (\PRIM\ 1\ (\CAR\  \pl))))
            \UNL{0}
	    \UNL{0}  (\DEFINE\ (\append\  \lista\ \listb)
	    \UNL{1}  (\SIF~(\NULLQ \ \lista)
	    \listb
	    \UNL{2} \hspace*{-0.1cm}(\CONS\ (\CAR\  \lista) (\append\
            (\CDR\  \lista)
            \listb))))
            \UNL{0}
            \UNL{0} (\LET\ \px\
            $\leftarrow$\ (\CONS\ $5$
            (\CONS\ (\CONS\ $6$ \NIL) \NIL) \IN
	    \UNL{1} (\LET\ \py\   $\leftarrow$\  (\CONS\ $3$ \NIL) \IN
	    \UNL{2}
            (\LET\ \pz\  $\leftarrow$\  (\append\ \px\  \py)\ \IN\
            \UNL{3} (\SIF~(\NULLQ~(\CAR~\pz))~$0$~$\pi$:\
            (\length\ \pz))))))
	  \end{uprogram}
      }}
      \\ 
      (a) Example program. \\  \\
%      \includegraphics[width=.45\textwidth]{motiv-example}
      \input{mem-graph} \\ 
      \renewcommand{\arraystretch}{.9}
      \begin{tabular}[t]{p{0.9\columnwidth}}
        (b) Memory graph at $\pi$.  \scalebox{.7}{\TwoCellsAD{a1}{a2}}
        denotes a closure. Thick edges denote live links. Traversal
        stops at edges marked $\times$ during garbage collection for
        an ideal collector.
      \end{tabular}
    \end{tabular} 
  \end{pspicture}
  \vspace*{-3ex}
  \caption{Example Program and its Memory Graph}\label{fig:mot-example}
\end{figure}

@@

The edges  shown by thick  arrows are those  which are live  at $\pi$.
Ideally, a cell should be  preserved during garbage collection only if
it is reachable from  the root set through a path  of live edges.  All
other  cells can  be reclaimed.   Thus if  a garbage  collection takes
place at $\pi$ with the heap shown in Figure~\ref{fig:mot-example}(b),
an ideal liveness-based  collector (LGC) will preserve  only the cells
referenced by  $\pz$, $(\CDR~ \pz)$  and (the cells  constituting) the
closure   referenced   by   $(\CDR~(\CDR~\pz))$.    In   contrast,   a
reachability-based collector (RGC) would have preserved all cells.

@@

@@

 Since  in a  lazy language,  an  expression is  not evaluated  unless
 required, the null demand ($\emptyset$) does not generate liveness in
 any  of  the  rules  defining  $\mathit{ref}$  or  $\mathcal{L}$.   A
 non-null  demand of  $\sigma$ on  (\CDR~$x$), is  transformed to  the
 liveness $\{x.\epsilon, x.\acdr\sigma\}$.  In  an opposite sense, the
 demand  of $\acdr\sigma$  on  (\CONS~$y$~$z$) is  transformed to  the
 demand  $\sigma$  on $z$.   Since  \CONS\  does not  dereference  its
 arguments, there  is no $\epsilon$ demand  on $y$ and $z$.  The rules
 for (\PRIM~x~y) and (\NULLQ~x) are similar. Constants do not generate
 any liveness.

@@

Section~\ref{sec:computing}   shows   how  the   demand   transformers
$\Lfonly$  for  a  program  (representing  a  fully  context-sensitive
analysis) can be safely approximated  by a {\em procedure summary} for
each function.   The summary is  in the  form of a  demand transformer
that maps a demand on a call to the function to demands on each of its
arguments.
